The Name of the Title is Hope

### 2023-06-19 input/2_introduction.tex
Updates from Overleaf

\section{Introduction}
% 1. Describe the Current Situtation
% Functions as a starting point and a common basis. Therefore it primarily contains recognizable and agreed points.
% 2. What is the complication, challenge identified
% Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.

Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities as well as simplifying skill acquisition~\citep{Kasahara2019-sk, Goto2020-mw}. There are two main technologies to physically augment users' actions: First, through the use of mechanical actuators, exoskeletons, a user's body can be moved by applying forces to the extremities, for example by pulling on the fingertips, and second, by making the user's extremities move by sending current into their muscle activating nerves using electrical muscle stimulation (EMS). However, augmented user's frequently report dissociative experiences during or following augmented action~\citep{}. They do not experience agency.
\todo{this needs a good citation}

The experience of agency means being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us. It has frequently been shown that if users have a strong sense of agency, they are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system~\citep{Berberian2012-do, Miller2007-rb}. The key challenge to drive adoption of human \textit{action augmentation} then is to design for agency experience, so user's feel as though they are in the ``driving seat'' once again.
% if something happens for you in an interaction there is a cost to that:
% - impacts sense of agency which in turn may makes user's engage less -> citation?
% - may impact precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control -> citation?

% 3. Question
% Asks the question how the hurdles of the Complication can be overcome. How can we prevent the threat or seize the opportunity? Also, what would be the benefits if the complication would be overcome?
% What have others done to address this and what do we propose "instead"
Previously, ~\citet{Kasahara2019-sk} have shown that reaction times when tapping targets on a touchscreen can be increased by augmenting the user through the use of EMS while maintaining their sense of agency. The authors found a relation between the experience of agency and the \textit{pre-emptive} gain, i.e. how much earlier the action of the user can be elicited. However, setting this gain factor may depend on each specific task and requires fitting the parameters over and over again. A key question then remains: How can we maintain agency during physical action augmentation without any training or running the risk of over-fitting task specific parameters? In other words, how can we design closed systems without the need for external tuning, to deliver an integrated, \textit{natural}, experience of user's augmented actions?
% At around 80 ms preceding the naturally occurring action, user's integrate the augmentation and maintain an experience of agency.

@@

% 4. (short) Answer Teaser
% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
% keep this short
In this paper, we present a fast communication channel between the user's brain signals and a physical end effector, here EMS. Crucially, our system controls the user's muscles through EMS at the time of the user's \textit{intent to act}, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG). In our user study we applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their sense of agency.
% We found ... rendering the augmentation a \textit{natural}, integrated expansion of the user's body.

>>>>

\section{Introduction}
% 1. Describe the Current Situtation
% Functions as a starting point and a common basis. Therefore it primarily contains recognizable and agreed points.
% 2. What is the complication, challenge identified
% Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.
\todo{add something even more general here: e.g. increase human wellbeing by XY}
Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities as well as simplifying skill acquisition~\citep{Kasahara2019-sk, Goto2020-mw}. There are two main technologies to physically augment users' actions: Through the use of mechanical actuators, i.e. exoskeletons, a user's body can be moved by applying forces to the extremities, for example by pulling on the fingertips. Another technology makes the user's extremities move by sending current into their muscle activating nerves using electrical muscle stimulation (EMS). However, augmented users often report dissociative experiences, frequently disrupting their sense of agency~\citep{Gilbert2017-ze, Gilbert2019-uc}.

Having a sense of agency means experiencing being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us. It has frequently been shown that if users have a strong sense of agency, they are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system~\citep{Berberian2012-do, Miller2007-rb}. The key challenge to drive adoption of human \textit{action augmentation} then is to design for agency experience, so users feel as though they are in the ``driver's seat'' once again.
% if something happens for you in an interaction there is a cost to that:
% - impacts sense of agency which in turn may makes user's engage less -> citation?
% - may impact precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control -> citation?

% 3. Question
% Asks the question how the hurdles of the Complication can be overcome. How can we prevent the threat or seize the opportunity? Also, what would be the benefits if the complication would be overcome?
% What have others done to address this and what do we propose "instead"
Previously, ~\citet{Kasahara2019-sk} have shown that reaction times when tapping a stimuli on a touchscreen can be reduced by augmenting the user with EMS while maintaining their sense of agency. The authors found a relation between the experience of agency and the \textit{pre-emptive} gain, i.e., how much earlier the action of the user could be elicited. However, setting this gain factor only works in stimulus-response scenarios where the user behavior can be predicted with high accuracy. A key question then remains: How can we maintain agency during physical action augmentation without controlling the environment? In other words, how can we design a closed-loop system to deliver an integrated, \textit{natural}, experience for users' augmented actions?
% At around 80 ms preceding the naturally occurring action, user's integrate the augmentation and maintain an experience of agency.

@@

% 4. (short) Answer Teaser
% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
% keep this short
In this paper, we present a fast communication channel between a user's brain signals and a physical end effector, here EMS. Crucially, our brain-computer interface controls the user's muscles through EMS at the time of their \textit{intent to act}, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG). In our user study we applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their sense of agency.
% We found ... rendering the augmentation a \textit{natural}, integrated expansion of the user's body.

### 2023-06-19 input/4_prototype.tex
Updates from Overleaf

\subsection{Controlling actuated haptic experiences using brain signals reflecting the intent to interact}

To discriminate between the two \textit{action} states, we leveraged the Readiness Potential (RP), or \textit{lateralized readiness potential}. It is an amplitude fluctuation that has been frequently observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}. It is reliably observed in electrodes placed over the motor cortex contralateral to the acting hand. Since its measurable onset precedes the time of participants self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective. However, evidence abounds for its role in action preparation. An RP is typically comprised of two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement. The first stage manifests in the pre-supplementary motor area and transfers to the premotor cortex shortly after. The second stage manifests contra-laterally in the primary motor cortex~\cite{Shibasaki2006-mt}. A recent study has shown that the RP is heavily ingrained in the subconscious mechanisms preceding movements that people cannot explicitly suppress~\cite{Schultze-Kraft2021-cu}. In their study,~\cite{Schultze-Kraft2021-cu} asked participants to find a way to perform a voluntary movement so that the accompanying RP amplitude was as small as possible. After each trial they informed participants about the strength of the RP in the current trial, so participants had a feedback metric to optimize for. They found participants unable to suppress their RP. This inability to suppress the RP renders it a reliable feature for classification.

% While its role in movement preparation remains somewhat controversial, it is established as a signal occurring during movement planning and motivates a usage as a reliable predictor of movement.

% \subsection{Action augmentation}
% TODO shorten this down to only the 'prototype' part -> EEG controlling the arduino switch. make obvious that what is attached to the switch can be different -> here in this particular study we used EMS -> see User Study

% \subsubsection{Apparatus}
% The experimental setup, depicted in Figure~\ref{}, comprised: (2) a 64-channel EEG system, and (3) a medically-compliant EMS device connected via two electrodes worn on the forearm. To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code to the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.

% \indent\textbf{(2) EEG Recording.} EEG data was recorded from 64 actively amplified Ag/AgCl electrodes in an actiCap Snap cap using BrainAmp DC amplifiers from BrainProducts. Electrodes were placed according to the extended international 10–20 system \cite{Chatrian1985-ys}. One electrode was placed under the left eye to provide additional information about eye movements (vEOG). After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity and electrode impedance was brought below 10k$\Omega$ for all electrodes. EEG data was recorded with a sampling rate of 250 Hz. We used LSL to make the EEG data stream available in the network and synchronize the recordings of EEG data, motion capture and an experiment marker stream that marked sections of the study procedure.

% \missingfigure{show participant with measurement system and description next to each element}

% \indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the index finger via electrical muscle stimulation (EMS), which was delivered via two electrodes attached to the participants' extensor digitorum muscle. We utilized the extensor digitorum since we found that we can robustly actuate it without inducing parasitical motion of neighboring muscles. This finger actuation was achieved via a medically-compliant battery powered muscle stimulator (prorelax TENS/EMS Super Duo Plus). The EMS system's output was controlled by flipping a solid state relay connected via an Arduino Uno to the experiment computer. The EMS was pre-calibrated per participant to ensure a pain-free stimulation and robust actuation.

\subsection{Training of Classifiers from Training Phase Data}
% Brain-Computer Interface to Detect EEG Readiness Potentials }
% Classifiying EEG Signals of the Intent to Interact

Before the test phase of our experiment, the data recorded during the training phase was processed in three consecutive steps: (1) we extracted movement onsets in each trial to construct three event classes: an \textit{idle} class representing resting background activity, a \textit{movement} class and a \textit{pre-movement} class, representing the intent to interact. Next, (2) a motion classifier was trained on \textit{movement} and \textit{idle} data segments and (3) an EEG classifier was trained on \textit{pre-movement} and \textit{idle} data segments using 10 features from the 20 most informative EEG channels. To classify single-trial motion and EEG data we used a regularized linear discriminant analysis (LDA) that was trained for each participant individually. The EEG classifier was then used in the test phase to detect RPs and trigger EMS. The motion classifier was used to make the interaction more robust. When it detected the participants hand to be moving, it deactivated the control loop from RP to EMS.

\missingfigure{some picture from the setup and prototype}

\indent\textbf{(1) Labelling Pre-movement, Movement and Idle Data Segments.}
In order to obtain training data segments for the \textit{pre-movement} class, a detection algorithm was applied on the hand velocity time series. From the training data, the raw hand motion data was filtered with a 6Hz low-pass filter and re-sampled to match the EEG sample rate using BeMoBIL Pipeline functions \textit{xdf2bids} and \textit{bids2set}\footnote{https://github.com/BeMoBIL/bemobil-pipeline}. Subsequently the first derivative was computed and velocity was extracted. Then, for each trial a two step thresholding process was applied to extract the exact time of movement onset: (1) the first time point when the velocity exceeded 70\% of the maximum velocity in the trial was selected, then (2) the signal was flipped at that time point and the time point where the flipped signal first fell below 10\% of the maximum in the remaining data was selected as the movement onset.

The three event classes were then defined as follows: \textit{pre-movement} from -1000 to 0 ms preceding the movement onsets, \textit{movement} from 0 to 1000 ms succeeding the movement onsets and \textit{idle}, a one second data segment that fell exactly between two succeeding trials, i.e. time-locked to the inter-trial-interval. Here, participants where looking at a fixation cross on the screen with their hand resting on the table.

\indent\textbf{(2) Training of Hand Motion Classifier.}
An LDA was trained using the mean velocity per trial and movement and idle time segment. Hence, 60 values for each class were subjected to train the motion classifier. Since in the idle time segments almost no movement occurred we added some artificial noise. To this end, the idle features were multiplied by a factor of 3, chosen by simple trial and error. In the real-time feedback, this rendered the classifier less susceptible to false positives, i.e. detecting a movement when there was none.

\indent\textbf{(3) Channel Selection \& Training of EEG Classifier.}
For the EEG classifier, we followed the approach outlined by Schultze-Kraft et al.\cite{} with slight modifications. First, the 20 channels with the strongest discrimination between \textit{pre-movement} and \textit{idle} segments were selected with the following procedure. For both \textit{pre-movement} and \textit{idle} the mean signal in the last 100 ms of each segment was subtracted from the mean signal in the first 100ms of the segment. The resulting value thus indicated how much the signal had changed in the course of the 1s segments. In line with the literature there should be a change over time in the \textit{pre-movement} segment but not in the \textit{idle} segment. Then, all channels were sorted (1) in descending order by the signal difference in the \textit{pre-movement} segments and (2) in ascending order by the signal difference in the \textit{idle} segments. The idea was, that an ideal channel shows a strong change in signal over the course of a \textit{pre-movement} segment but shows no difference during a \textit{idle} segment. The ranks were joined and the 20 best channels were selected. In any case, channels C3, C4 and Cz were kept for every participant as these are most frequently reported in studies on the RP.

To train the LDA EEG classifier, we extracted spatiotemporal features from the EEG; 10 windowed means were extracted for each segment and channel. These were baseline corrected by subtracting the mean of the last 100 ms preceding the segment. The features were concatenated across all 20 selected channels and 10 time windows to obtain a 200 dimensional spatiotemporal feature vector per \textit{pre-movement} and \textit{idle} segment.

\subsubsection{Real-time Feedback}

During the test phase, both classifiers were applied to the real-time data. With the motion data streaming at 90 Hz we subjected the current sample to the motion classifier. For the EEG data streaming at 250 Hz, the EEG data of the last 1000ms was buffered for the selected best channels and 10 windowed means per channel were computed. Identical to the training data, these features were baseline corrected by subtracting the mean from the last 100 ms preceding the last second, i.e. -1100 ms to -1000 ms, resulting in a 200 dimensional baseline corrected spatiotemporal feature vector. Hence, both classifiers yielded one output value at each sample point.

The EMS was triggered whenever the EEG classifier predicted the \textit{pre-movement} class with an above .7 probability. Additionally, EMS was only triggered when the hand was \textit{not} moving. In each trial, as soon as the hand was in motion once after trial start, the EEG-EMS control was deactivated for the rest of the trial. This guaranteed that participants experienced EMS only during movement preparation.

>>>>

\section{Agency Prototype}
Here we present a prototype to overcome the disruption in experiencing agency during action augmentation. Our proposed system controls an end-effector, here EMS, by leveraging EEG signals reflecting the intent to (inter)act. To control the interaction in real-time, an EEG classifier is used to discriminate between two user states: either the user is predicted to be in an \textit{idle} state, reflecting the absence of an intent to act-- or a \textit{pre-movement} state, indicating the presence of an intent to act. When the system predicts a \textit{pre-movement} state, EMS is triggered in real-time, augmenting the user's action even before their own voluntary motor command.

\subsection{Controlling actuated haptic experiences using brain signals reflecting the intent to interact}
% alternative headings:
% Triggering EMS using EEG Signals

To discriminate between the two \textit{action} states, we leveraged the EEG Readiness Potential (RP), or \textit{lateralized readiness potential}. It is an amplitude fluctuation that has frequently been observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}. The RP is reliably observed at electrodes placed over the motor cortex contralateral to the acting hand. Since its measurable onset precedes the time of participants self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective. However, evidence abounds for its role in action preparation. A RP is typically comprised of two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement. The first stage manifests in the pre-supplementary motor area and transfers to the premotor cortex shortly after. The second stage manifests contra-laterally in the primary motor cortex~\cite{Shibasaki2006-mt}. A recent study has shown that the RP is heavily ingrained in the subconscious mechanisms preceding movements that people cannot explicitly suppress~\cite{Schultze-Kraft2021-cu}. In their study,~\cite{Schultze-Kraft2021-cu} asked participants to find a way to perform voluntary movements while keeping accompanying RP amplitudes as small as possible. After each trial they informed participants about the strength of the RP in the current trial, so participants had a feedback metric to optimize for. They found participants unable to suppress their RP. This inability to suppress the RP renders it a reliable feature for classification.

\subsection{Prototype components}
Our system, depicted in Figure~\ref{}, comprised: (1) a 64-channel EEG system, (2) a 1 channel Electromyography (EMG) system, (3) a simple Arduino controlled switch, and (4) a medically-compliant EMS device connected via two electrodes worn on the arm. More information on the specific hardware specifications is given in the methods and user study sections below. To assist readers in replicating our prototype, we provide the necessary technical details and source code for data processing, classifier training, and real-time application\footnote{[anonymized]}. We used LSL\footnote{https://github.com/sccn/labstreaminglayer} to make the EEG and EMG data streams available in the network and synchronize the recordings of EEG data, motion capture, and an experiment marker stream that marked sections of the study procedure.

\missingfigure{show participant with measurement system and description next to each element}

\subsection{Brain-Computer Interface}
% alternative headings:
% Brain-Computer Interface to Detect EEG Readiness Potentials }
% Classifiying EEG Signals of the Intent to Interact

The presented classification scheme required individual training data per participant, as transfer learning remains challenging with EEG~\cite{Wan2021-zz}. Therefore, we initially obtained training data on a variant of our task that did not include any muscle stimulation. In our task, participants were instructed, following an initial waiting period, to tap a touchscreen at their own volition with their ring finger, see figure ~\ref{fig:teaser}.

\subsubsection{Labelling movement classes using EMG}
To obtain behavioral labels at a high temporal resolution, we leveraged EMG from the flexor digitorum profundus to detect finger movement onsets of the ring finger. Subsequently the data were labelled into two classes: an \textit{idle} class representing resting background activity and a \textit{pre-movement} class, representing the intent to (inter-)act. To this end, EMG amplitudes of the second preceding the screen tap were band-pass filtered from 20 to 100 Hz and subsequently squared. Next, noisy data segments were determined. A segment was rejected as an outlier using Matlab's \textit{isoutlier} function using default parameters. To label the time of movement onset, the first sample where the amplitude exceeded the 95th percentile of the data epoch was selected. The two event classes were then defined as follows: \textit{pre-movement} from -1000 to 0 ms preceding the movement onsets and \textit{idle}, a one second data segment between trials where participants were looking at a fixation cross.
% from isoutlier documentation (mean segment amplitude exceeding 3 scaled median absolute deviation)

% \missingfigure{some picture from the setup and prototype}

\subsubsection{Selecting discriminative EEG channels}
In a first step to prepare the EEG data for classification, noisy data segments were rejected using the identical method as for rejection of EMG segments described above. Then, to decrease dimensionality of the EEG data, a selection of discriminative channels was conducted following an approach outlined by~\citep{Schultze-Kraft2021-cu} albeit with slight modifications.

First, for each channel and both \textit{idle} and \textit{pre-movement} class, the mean slope of the amplitude was obtained by fitting a linear regression, indicating how much the signal had changed in the course of the 1s segments. In line with the literature on the RP, there should be a change over time in the \textit{pre-movement} segment but not in the \textit{idle} segment. Then, all channels were sorted (1) in descending order by slope in the \textit{pre-movement} segments and (2) in ascending order by slope in the \textit{idle} segments. The idea was, that an ideal channel shows a strong change in signal over the course of a \textit{pre-movement} segment but shows no difference during a \textit{idle} segment. The ranks were joined and the 10 highest-ranking channels were selected. In any case, channels C3, C4, and Cz were included in the 10 selected channels for every participant even if they were not within the 10 highest ranks as these are most frequently reported in studies on the RP.

\subsubsection{Training of EEG classifier}
To classify EEG data a regularized linear discriminant analysis (LDA) was trained for each participant individually. The EEG classifier was then used in the test phase to detect RPs and trigger EMS. The classifier was trained using the slopes of the \textit{idle} and \textit{pre-movement} segments of the 10 most discriminative channels as features, hence comprising a 10 dimensional feature vector.
\todo{extend this after settling on a final feature set}

\subsubsection{Real-time application and EMS control}
During real-time application, the EEG data was buffered for the last second for the selected channels. The slope features were then computed and applied to the model. An Arduino was used to flip a switch whenever the EEG classifier predicted the \textit{pre-movement} class with an above .7 probability, thereby triggering the attached actuating hardware. 
\todo{this may still change, adapt accordingly later}
In our current prototype, we used a medically-compliant EMS device. We constrained the times when the switch could be flipped to specific moments in the experiment where participants were getting ready to perform an action, thereby limiting the impact of false-positive classification. 
% This guaranteed that participants experienced EMS only during movement preparation.

%%%% Resources

% old version
% First, the 10 channels with the strongest discrimination between \textit{pre-movement} and \textit{idle} segments were selected with the following procedure: For both \textit{pre-movement} and \textit{idle} the mean signal in the last 100 ms of each segment was subtracted from the mean signal in the first 100ms of the segment. The resulting value thus indicated how much the signal had changed in the course of the 1s segments. In line with the literature there should be a change over time in the \textit{pre-movement} segment but not in the \textit{idle} segment. Then, all channels were sorted (1) in descending order by the signal difference in the \textit{pre-movement} segments and (2) in ascending order by the signal difference in the \textit{idle} segments. The idea was, that an ideal channel shows a strong change in signal over the course of a \textit{pre-movement} segment but shows no difference during a \textit{idle} segment. The ranks were joined and the 10 best channels were selected. In any case, channels C3, C4 and Cz were kept for every participant as these are most frequently reported in studies on the RP.
% An EEG classifier was trained on \textit{pre-movement} and \textit{idle} data segments using 10 features from the 20 most informative EEG channels. To classify single-trial motion and EEG data we used a regularized linear discriminant analysis (LDA) that was trained for each participant individually. 
% These were baseline corrected by subtracting the mean of the last 100 ms preceding the segment. The features were concatenated across all 20 selected channels and 10 time windows to obtain a 200 dimensional spatiotemporal feature vector per \textit{pre-movement} and \textit{idle} segment.

% During the test phase, both classifiers were applied to the real-time data. With the motion data streaming at 90 Hz we subjected the current sample to the motion classifier. For the EEG data streaming at 250 Hz, the EEG data of the last 1000ms was buffered for the selected best channels and 10 windowed means per channel were computed. Identical to the training data, these features were baseline corrected by subtracting the mean from the last 100 ms preceding the last second, i.e. -1100 ms to -1000 ms, resulting in a 200 dimensional baseline corrected spatiotemporal feature vector. Hence, both classifiers yielded one output value at each sample point.
